\section{Common Java packages}\label{appendix:packages}

Descriptions from: \url{https://docs.oracle.com/javase/8/docs/api/java/util/package-summary.html}.

TODO: What about interfaces such as List or other implementations such as LinkedList (vs. ArrayList)? What about more advanced data structures like maps? More granular types like \ic{java.lang.Long}?

\begin{fullwidth}
\begin{center}
\begin{tabularx}{\linewidth}{ l X }
\ic{java.lang.*} & Provides classes that are fundamental to the design of the Java programming language.
\\\\
\ic{java.lang.Object} & Class \ic{Object} is the root of the class hierarchy. Every class has \ic{Object} as a superclass. All objects, including arrays, implement the methods of this class.
\\\\
\ic{java.lang.Double} & The \ic{Double} class wraps a value of the primitive type \ic{double} in an object.
\\\\
\ic{java.lang.Integer} & The \ic{Integer} class wraps a value of the primitive type \ic{int} in an object.
\\\\
\ic{java.lang.Long} & The \ic{Long} class wraps a value of the primitive type \ic{long} in an object.
\\\\
\ic{java.lang.Math} & The class \ic{Math} contains methods for performing basic numeric operations such as the elementary exponential, logarithm, square root, and trigonometric functions.
\\\\
\ic{java.lang.Random} & An instance of this class is used to generate a stream of pseudorandom numbers.
\\\\
\ic{java.lang.String} & The String class represents character strings.
\\\\
\ic{java.lang.System} & The \ic{System} class contains several useful class fields and methods, most notably the \ic{out} field's \ic{println} method. It cannot be instantiated.
\\
\hline
\ic{java.util.*} & Contains the collections framework, legacy collection classes, event model, date and time facilities, internationalization, and miscellaneous utility classes (a string tokenizer, a random-number generator, and a bit array).
\\\\
\ic{java.lang.ArrayList} & An ordered collection (also known as a \emph{sequence}).
\\\\
\ic{java.util.Scanner} & A simple text scanner which can parse primitive types and strings using regular expressions.
\\
\hline
\ic{java.io} & \text{}
\end{tabularx}
\end{center}
\end{fullwidth}